# Contact Prior {#contact-prior}

The wealth of successful meta-predictors presented in section \@ref(meta-predictors) highlights the importance to exploit other sources of information apart from coevolution statistics.
Much information about residue interactions is typically contained in features of 1D properties at positions $i$ and $j$ predicted from local sequence profiles, such as secondary structure, solvent accessibility or contact number, and in features of predicted 2D properties such as the contact prediction scores for $(i,j)$ from a profile-based method. 

For example, predictions of secondary structure elements and solvent accessibility are used by almost all modern machine learning predictors, such as MetaPsicov [@Jones2015a], NeBCon [@He2017], EPSILON-CP [@Stahl2017], PconsC3 [@Skwark2016].
Other frequently used sequence derived features include pairwise contact potentials, sequence separation and conservation measures such as column entropy [@Jones2015a; @He2017; @Ma2015a].

In the following sections I present a random forest classifier that uses sequence derived features to distinguish contacts from non-contacts.
Methods section \@ref(seq-features) lists all features used to train the classifier including the aforementioned standard features as well as some novel features. 

The probabilistic predictions of the random forest model can be introduced directly as prior information into the Bayesian statistical model presented in the last section \@ref(bayesian-approach) to improve the overall prediction accuracy in terms of posterior probabilities.
Furthermore, contact scores from coevolution methods can be added as an additional feature to the random forest model in order to elucidate how much the combined information improves prediction accuracy over the single methods. 


## Random Forest Classifiers

Random Forests are supervised machine learning methods that belong to the class of ensemble methods [@Ho1998;@TinKamHo;@Breiman2001].
They are easy to implement, fast to train and can handle large numbers of features due to implicit feature selection [@Menze2009].

Ensemble methods combine the predictions of several independent base estimators with the goal to improve generalizability over a single estimator.
Random forests are ensembles of decision trees where randomness is introduced in two ways:

1. every tree is build on a random sample that is of the same size but drawn with replacement from the training set (i.e., a bootstrap sample) 
2. every split of a node is performed on a random subset of features

Predictions from all trees are averaged to obtain the final prediction.

A single decision tree, especially when it is grown very deep is highly susceptible to noise in the training set and therefore prone to overfitting which results in poor generalization ability. 
As a consequence of randomness and averaging over many decision trees, the variance of a random forest predictor decreases and therefore the risk of overfitting.

Random forests are capable of regression and classification tasks.
For classification, predictions for new data are obtained by running a data sample down every tree in the forest and then either apply majority voting over single class votes or averaging the probabilistic class predictions. Probabilistic class predictions of single trees are computed as the fraction of training set samples of the same class in a leaf whereas the single class vote refers to the majority class in a leaf. 

Typically, *Gini impurity*, which is a computationally efficient approximation to the entropy, is used as a split criterion to estimate the quality of a split of a node.
It measures the degree of purity in a data set regarding class labels as $GI = (1 - \sum_{k=1}^K p_k^2)$, where $p_k$ is the proportion of class $k$ in the data set.
The feature $f$ with the highest *decrease in Gini impurity* $\Delta GI_f$ over the two resulting child node subsets will be used to split the data set at the given node $N$, 

$$
\Delta GI_f(N_{\textrm{parent}}) = GI_f(N_{\textrm{parent}}) - p_{\textrm{left}} GI(N_{\textrm{left}}) - p_{\textrm{right}} GI(N_{\textrm{left}})
$$

where $p_{\textrm{left}}$ and $p_{\textrm{right}}$ refers to the fraction of samples ending up in the left and right child node respectively [@Menze2009].


Summing the *decrease in Gini impurity* for a feature $f$ over all trees whenever $f$ was used for a split yields the *Gini importance* measure, which can be used as an estimate of general feature relevance.
Random forests therefore are popular methods for feature selection and it is common practice to remove the least important features from a data set to reduce the complexity of the model.
However, feature importance measured with respect to *Gini importance* needs to be interpreted with care.
The random forest model cannot distinguish between correlated features and it will choose any of the correlated features for a split, thereby reducing the importance of the other features and introducing bias.
Furthermore, it has been found that feature selection based on *Gini importance* is biased towards selecting features with more categories as they will be chosen more often for splits and therefore tend to obtain higher scores [@Strobl2007].


## Evaluating Random Forest Model as Contact Predictor

I trained a random forest classifier on the feature set described in methods section \@ref(seq-features) and optimized model hyperparameters as well as some data set specific settings (e.g window size and class ratios) with 5-fold cross-validation as described in methods section \@ref(rf-hyperparameter-optimization).

Figure \@ref(fig:rf-feature-importance) shows the ranking of the ten most important features according to *Gini importance*.
Both local statistical contact scores, *OMES* [@Fodor2004a] and [MI](#abbrev) (mutual information between amino acid counts), constitute the most important features besides the mean pair potentials acording to Miyazawa & Jernigan [@Miyazawa1999a] and Li&Fang[@Li2011].
Further important features include the relative solvent accessibility at both pair positions, the total percentage of gaps at both positions, the correlation between mean isoelectric point property at both positions, sequence separation and the beta-sheet propensity in a window of size five around position i. 

(ref:caption-rf-feature-importance) Top ten features ranked according to *Gini importance*. Features are described in detail in methods section \@ref(seq-features).

```{r rf-feature-importance, echo = FALSE, screenshot.alt="img/random_forest_contact_prior/feature_random_forest_optimalhyperparameters_topfeatures.png", out.width = '100%', fig.cap = '(ref:caption-rf-feature-importance)'}
knitr::include_url("img/random_forest_contact_prior/feature_random_forest_optimalhyperparameters_topfeatures.html", height = "500px")
```

Many features have low *Gini importance* scores which means they are rarely considered for splitting a node and can likely be removed from the dataset. 
Removing irrelevant features from the dataset is a convenient procedure to reduce model complexity.
As described in methods section \@ref(rf-feature-selection), I performed feature selection by evaluating model performance on subsets of features of decreasing importance.
Most models trained on subsets of the total feature space perform nearly identical compared to the model trained on all features, as can be seen in Figure \@ref(fig:rf-performance-feature-selection).
When using only the 26 most relevant features, performance of the model drops compared to using at least 75 features.
The model using the 75 most important features has been selected for further analysis. 

(ref:caption-rf-feature-selection) Average precision of random forest models trained on subsets of sequence derived features. Subsets of features have been selected as described in methods section \@ref(rf-feature-selection).  

```{r rf-feature-selection, echo = FALSE, screenshot.alt="img/random_forest_contact_prior/precision_vs_rank_featureselection_random_forest_optimized_hyperparameters.png", out.width = '100%', fig.cap = '(ref:caption-rf-feature-selection)'}
knitr::include_url("img/random_forest_contact_prior/precision_vs_rank_featureselection_random_forest_optimized_hyperparameters.html", height = "500px")
```


Figure \@ref(performance-rf) shows the mean precision for the random forest model trained on the 75 most important feautures and for several contact scores.
The pseudo-likelihood contact score ([APC](#abbrev) corrected Frobenius norm) is shown as a reference method.
The random forest model has a mean precision of 0.33 for the top 5L contacts compared to a precision of 0.47 for pseudo-likelihood.
Furthermore, the random forest model improves approximately ten percentage points in precision over the local statistical contact scores, *OMES* and mutual information (MI).
Both methods comprise important features of the random forest model as can be seen in Figure \@ref(fig:rf-feature-importance).


(ref:caption-performance-rf) Mean precision for top ranked contacts on a test set of 761 proteins. **omes_fodoraldrich+apc** = OMES score with APC as described in section \@ref(seq-features-pairwise). **mi_pc + APC** = mutual information with APC as described in section \@ref(seq-features-pairwise). **rf_contact_prior** = random forest model using only sequence derived features.  **pLL-L2normapc-RF** = random forest model using sequence derived features and pseudo-likelihood contact score (L2norm + APC). **ccmpred-pll-centerv+apc** = conventional pseudo-likelihood contact score (L2norm + APC)

```{r performance-rf, echo = FALSE, screenshot.alt="img/random_forest_contact_prior/precision_vs_rank_notitle.png", out.width = '100%', fig.cap = '(ref:caption-performance-rf)'}
knitr::include_url("img/random_forest_contact_prior/precision_vs_rank_notitle.html", height = "500px")
```


When analysing performance with respect to alignment size it can be found that the random forest model outperforms the pseudo-likelihood score for small alignments (see Figure \@ref(fig:performance-neff-rf)).  
Both, local statistial models *OMES* and [MI](#abbrev) also perform weak on small alignments, leading to the conclusion that the remaining sequence derived features are highly relevant when the alignment contains only few sequences.
This finding is expected, as it is well known that models trained on simple sequence features perform almost independent of alignment size [@Stahl2017, @Skwark2016].


(ref:caption-performance-neff-rf) Mean precision for top ranked contacts on a test set of 761 proteins splitted into four equally sized subsets according to [Neff](#abbrev). Subsets are defined according to quantiles of [Neff](#abbrev) values. Upper left: Subset of proteins with [Neff](#abbrev) < Q1. Upper right: Subset of proteins with Q1 <= [Neff](#abbrev) < Q2. Lower left: Subset of proteins with Q2 <= [Neff](#abbrev) < Q3. Lower right: Subset of proteins with Q3 <= [Neff](#abbrev) < Q4. **omes_fodoraldrich+apc** = OMES score with [APC](#abbrev) as described in section \@ref(seq-features-pairwise). **mi_pc + APC** = mutual information with [APC](#abbrev) as described in section \@ref(seq-features-pairwise). **rf_contact_prior** = random forest model using only sequence derived features.  **pLL-L2normapc-RF** = random forest model using sequence derived features and pseudo-likelihood contact score ([APC](#abbrev) corrected Frobenius norm of the couplings $\wij$). **ccmpred-pll-centerv+apc** = [APC](#abbrev) corrected Frobenius norm of the couplings $\wij$ computed with pseudo-likelihood.


```{r performance-neff-rf, echo = FALSE, screenshot.alt="img/random_forest_contact_prior/precision_vs_rank_facetted_by_neff_notitle.png", out.width = '100%', fig.cap = '(ref:caption-performance-neff-rf)'}
knitr::include_url("img/random_forest_contact_prior/precision_vs_rank_facetted_by_neff_notitle.html", height = "600px")
```



## Additional Coevolution Features improves Precision

Figure \@ref(fig:performance-neff-rf) showed that the random forest predictor improves over the pseudo-likelihood coevolution method when the alignment consists of only few sequences. 
In order to assess this improvement in a more direct manner, it is possible to build a combined random forest predictor that is not only trained on the sequence derived features but also on the pseudo-likelihood contact score as an additional feature (see methods section \@ref(rf-with-pll-score) for details). 
As expected, the pseudo-likelihood score comprises the most important feature in the model (Figure \@ref(fig:feature-importance-rf-with-pll-score)).

Finally, the comparison of the extended random forest model ("pLL-L2normapc-RF") to the pseudo-likelihood score in Figure \@ref(fig:performance-rf-with-pll-score) reveals that the additional coevolutionary feature indeed improves performance over the score without prior information.
Especially for small alignments, the improvment is substantial as can be seen in Figure \@ref(fig:performance-neff-rf-with-pll-score).
In contrast, the improvement on large alignments is small, as the gain from simple sequence features compared to the much more powerful coevolution signals is neglectable. 













